---
title: "Bakery Sales Mock Project"
author: "Anthony Morciglio"
date: "2024-07-25"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# Load libraries of the data:

library(ggplot2)
library(dplyr)
library(tidyverse)
library(readr)
library(skimr)
library(readxl)
library(viridis)
library(flextable)

options(warn = -1)

f1 <- "C:/Users/Anthony Morciglio/OneDrive/Coursera Google Data Analytics/Bakery Sales 2020 RW.csv"
df <- read.csv(f1)

```


## Slide with Bullets

This mock project consists of data points with different features such as the following:
- Name of the item (Nombre).
- Cost of the item (Cuesta).
- Quantity of the item (Cantidad).
- Revenue of the items (Dinero) calculated by taking the product of quantity and cost.

## Slide with R Output

```{r, echo = TRUE}

# Summarize key features of the data:
df
glimpse(df)
str(df)
summary(df)
skim_without_charts(df)

```

## Slide with Summarized Table

We can aggregate the data based on name type.  That is, group by name and summarize the total price and number of goods sold by using the group_by() and summarise() functions.

```{r, echo = TRUE}
TP <- df %>% group_by(Nombre) %>% summarise(Total_Price = sum(Cuesta * Cantidad), Total_Sold = sum(Cantidad)) %>% flextable()

```


## Slide with Graphing

```{r, echo = TRUE}
ggplot(data = TP, aes(x = Nombre, y = Total_Price), fill = Nombre) + 
  geom_bar(stat = identity) + 
    labs(x = "Nombre", y = "Total Price", title = "Bar Chart: Total Price vs Nombre", fill = "Nombre") +
  scale_fill_manual(values = viridis(n = length(unique(TP$Nombre)))) + 
  theme_linedraw(base_size = 18
                 
ggplot(data = TP, aes(x = Nombre, y = Total_Sold), fill = Nombre) + 
  geom_bar(stat = identity) + 
    labs(x = "Nombre", y = "Total Sold", title = "Bar Chart: Total Sold vs Nombre", fill = "Nombre") +
  scale_fill_manual(values = viridis(n = length(unique(TP$Nombre)))) + 
  theme_linedraw(base_size = 18

```
