---
title: "Central Limit Theorem with Binomial Example"
author: Anthony Morciglio
date: 10/02/2025
output: html_document
---

```{r, echo=TRUE, message=FALSE, warning=FALSE}
```

## Central Limit Theorem with the Binomial Distribution

We demonstrate how the Central Limit Theorem (CLT) approximates the Binomial distribution to the Normal distribution for a large number of samples.

### Example Setup

Suppose each light bulb has a probability $p = 0.1$ of being defective, and we test $n = 100$ bulbs. 
Then:

* Mean: $\mu = np = 10$
* Variance: $\sigma^2 = np(1-p) = 9$, so $\sigma = 3$

```{r, echo=TRUE, message=FALSE, warning=FALSE}
n <- 100
p <- 0.1
mu <- n * p
sigma <- sqrt(n * p * (1 - p))
mu; sigma
```

### Exact Binomial Probability

We compute the probability that at most 8 bulbs are defective.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
p_exact <- pbinom(8, size = n, prob = p)
p_exact
```

### Normal Approximation (with Continuity Correction)

```{r, echo=TRUE, message=FALSE, warning=FALSE}
z <- (8.5 - mu) / sigma
p_approx <- pnorm(z)
p_approx
```

### Monte Carlo Simulation

We use simulation to approximate $\mathbb{P}(Y \leq 8)$.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
set.seed(123)
sim <- rbinom(1e5, size = n, prob = p)
mean(sim <= 8)
```

### Visualization

We compare the binomial probabilities with the normal approximation.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)

# Range around the mean
y_vals <- 0:30
binom_probs <- dbinom(y_vals, size = n, prob = p)
norm_approx <- dnorm(y_vals, mean = mu, sd = sigma)

df <- data.frame(
  y = y_vals,
  Binomial = binom_probs,
  NormalApprox = norm_approx
)

ggplot(df, aes(x = y)) +
  geom_col(aes(y = Binomial), fill = "steelblue", alpha = 0.6) +
  geom_line(aes(y = NormalApprox), color = "black", size = 1.2) +
  labs(
    title = "Binomial Distribution vs Normal Approximation",
    y = "Probability",
    x = "Number of defective bulbs"
  ) +
  theme_minimal()
```

This shows how the CLT allows us to approximate the Binomial distribution with a Normal distribution when the number of trials is large (here $n=100$).


