---
title: "PBJ Analysis Q1 24"
author: "Anthony Morciglio"
date: "2024-10-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## CMS Data Exercise Part 1

This is an exercise for an interview for the company of Clipboard Health, dataset can be downloaded at the site: [Data.CMS.gov](https://data.cms.gov/quality-of-care/payroll-based-journal-daily-nurse-staffing/data?query=%7B%22filters%22%3A%7B%22rootConjunction%22%3A%7B%22label%22%3A%22And%22%2C%22value%22%3A%22AND%22%7D%2C%22list%22%3A%5B%5D%7D%2C%22keywords%22%3A%22%22%2C%22offset%22%3A10%2C%22limit%22%3A10%2C%22sort%22%3A%7B%22sortBy%22%3Anull%2C%22sortOrder%22%3Anull%7D%2C%22columns%22%3A%5B%5D%7D)

We can observe that the dataset consists of attributes such as provider, city, state, county, and hours worked for the occupations of CNA, LPN, and RN.  For this study, we will focus on the first quarter of 2024.  We are interested in exploring the measures of central tendency and measures of spread based on the geography.  Theoretically, we could explore how hours worked depends on the demographics of the population including gender, age, and nationality, but for simplicity, we will focus on geography.  

```{r, echo = FALSE}

# PBJ Daily Nurse Staffing Q12024 Analysis
## -- S1: Load packages -- ##
library(ggplot2)
library(dplyr)
library(tidyverse)
library(readr)
library(skimr)
library(readxl)
library(viridis)
library(flextable)

## -- S2: Load, Clean, and Extrapolate the Data -- ##
# f1 is revised to include population density and f2 is somewhat raw.
f1 <- "C:/Users/Anthony Morciglio/OneDrive/Projects/Programming/R/PBJ_Daily_Nurse_Staffing_Q1_2024_Revised.csv"
f2 <- "C:/Users/Anthony Morciglio/OneDrive/Projects/Programming/R/PBJ_Daily_Nurse_Staffing_Q1_2024_new.csv"

# Read the data that was cleaned in Excel and exported as a CSV file.
df <- read_csv(f1)
df2 <- read_csv(f2)

# Glance at the data
tail(df)
glimpse(df)
str(df)
summary(df) # summarize numeric columns
df %>% select_if(is.numeric) %>% skim() %>% select(-n_missing) 
# check for completeness of data



## -- Grouping Data -- ##
# Group data based on State and Provider
print("--- Summarizing key statistics based on State ---")
df_state <- df %>%  group_by(State) %>% 
  drop_na() %>% 
  summarize(avg_RN_Dir_hours = mean(Hrs_RNDON), z_RN_hours = (Hrs_RN - mean(Hrs_RN))/sqrt(var(Hrs_RN)), avg_RN_hours = mean(Hrs_RN), avg_RN_density = mean(Hrs_RN_Density), z_LPN_hours = (Hrs_LPN - mean(Hrs_LPN))/sqrt(var(Hrs_LPN)), avg_LPN_hours = mean(Hrs_LPN), avg_LPN_density = mean(Hrs_LPN_Density), z_CNA_hours = (Hrs_CNA - mean(Hrs_CNA))/sqrt(var(Hrs_CNA)), avg_CNA_hours = mean(Hrs_CNA), avg_CNA_Density = mean(Hrs_CNA_Density))

df_state_2 <- df2 %>%  group_by(State) %>% 
  drop_na() %>% 
  summarize(avg_RN_Dir_hours = mean(Hrs_RNDON), z_RN_hours = (Hrs_RN - mean(Hrs_RN))/sqrt(var(Hrs_RN)), avg_RN_hours = mean(Hrs_RN), z_LPN_hours = (Hrs_LPN - mean(Hrs_LPN))/sqrt(var(Hrs_LPN)), avg_LPN_hours = mean(Hrs_LPN), z_CNA_hours = (Hrs_CNA - mean(Hrs_CNA))/sqrt(var(Hrs_CNA)), avg_CNA_hours = mean(Hrs_CNA))


print("--- Summarizing key statistics based on Provider ---")
df_provider <- df %>%  group_by(Provname) %>% 
  drop_na() %>% 
  summarize(avg_RN_Dir_hours = mean(Hrs_RNDON), z_RN_hours = (Hrs_RN - mean(Hrs_RN))/sqrt(var(Hrs_RN)), avg_RN_hours = mean(Hrs_RN), avg_RN_density = mean(Hrs_RN_Density), z_LPN_hours = (Hrs_LPN - mean(Hrs_LPN))/sqrt(var(Hrs_LPN)), avg_LPN_hours = mean(Hrs_LPN), avg_LPN_density = mean(Hrs_LPN_Density), z_CNA_hours = (Hrs_CNA - mean(Hrs_CNA))/sqrt(var(Hrs_CNA)), avg_CNA_hours = mean(Hrs_CNA), avg_CNA_Density = mean(Hrs_CNA_Density))

df_provider_2 <- df2 %>%  group_by(Provname) %>% 
  drop_na() %>% 
  summarize(avg_RN_Dir_hours = mean(Hrs_RNDON), z_RN_hours = (Hrs_RN - mean(Hrs_RN))/sqrt(var(Hrs_RN)), avg_RN_hours = mean(Hrs_RN), z_LPN_hours = (Hrs_LPN - mean(Hrs_LPN))/sqrt(var(Hrs_LPN)), avg_LPN_hours = mean(Hrs_LPN), z_CNA_hours = (Hrs_CNA - mean(Hrs_CNA))/sqrt(var(Hrs_CNA)), avg_CNA_hours = mean(Hrs_CNA))


# Filter based on state of Georgia and Florida (just an example)
# Increase granularity of the data to reduce variability
df_state_ga <- filter(df, df$State == "GA")
df_state_fl <- filter(df, df$State == "FL")

df_state_ga %>% select_if(is.numeric) %>% skim() %>% select(-n_missing)
df_state_fl %>% select_if(is.numeric) %>% skim() %>% select(-n_missing)

# Filter based on County within same State - increase granularity
df_ga_fulton <- filter(df_state_ga, df_state_ga$County_Name == "Fulton")
df_ga_gwinnett <- filter(df_state_ga, df_state_ga$County_Name == "Gwinnett")

skim(df_ga_fulton) %>% flextable()
skim(df_ga_gwinnett) %>% flextable()

df_ga_fulton %>% select_if(is.numeric) %>% skim() %>% select(-n_missing)
df_ga_gwinnett %>% select_if(is.numeric) %>% skim() %>% select(-n_missing)


```

## Hypothesis Testing

We ask if there is a statistical difference between the population averages of hours worked by Licensed Practical Nurses (LPN's) between the states of Florida and Georgia.  Using a two sided T-test with a significance factor of 5%, we observe that there is no difference in population averages, but there is a statistical difference in the variances.  This suggests that there is a greater variability in the hours worked by LPN's (see density and histogram plots for FL vs GA).  We achieve similar results when comparing hours worked by Registered Nurses (RN's) in Gwinnett and Fulton counties (from Georgia).  


```{r Hypothesis Testing, echo=FALSE}

# Apply hypothesis testing for average and variance between states and counties

Fstat_gafl <- var.test(df_state_ga$Hrs_LPN, df_state_fl$Hrs_LPN, alternative = "two.sided")
var_diff_bool <- FALSE
tmp_df <- df %>% filter(df$State == "FL" | df$State == "GA")
Tstat_gafl <- t.test(tmp_df$Hrs_LPN ~ tmp_df$State, data = tmp_df, var.equal = !var_diff_bool)

if (Tstat_gafl[3]$p.value <= 0.05)
{ print("There is a Statistical difference between the average hours worked by LPN's from GA and FL") 
} else 
{ print("There is no Statistical difference between the average hours worked by LPN's from GA and FL")
}

if (Fstat_gafl[3]$p.value <= 0.05)
{ print("There is a Statistical difference between the variances of hours worked by LPN's from GA and FL") 
} else 
{ print("There is no Statistical difference between the variances of hours worked by LPN's from GA and FL")
}


# -- Apply same method for RN's between states -- #
Fstat_gwfu <- var.test(df_ga_fulton$Hrs_RN, df_ga_gwinnett$Hrs_RN, alternative = "two.sided")
var_diff_bool <- FALSE
tmp_df <- df_state_ga %>% filter(df_state_ga$County_Name == "Gwinnett" | df_state_ga$County_Name == "Fulton")
Tstat_gwfu <- t.test(tmp_df$Hrs_RN ~ tmp_df$County_Name, data = tmp_df, var.equal = !var_diff_bool)

if (Tstat_gwfu[3]$p.value <= 0.05)
{ print("There is a Statistical difference between the average hours worked by RN's in Gwinnett and Fulton Counties") 
} else 
{ print("There is no Statistical difference between the average hours worked by RN's in Gwinnett and Fulton Counties")
}

if (Fstat_gwfu[3]$p.value <= 0.05)
{ print("There is a Statistical difference between the variances of hours worked by in Gwinnett and Fulton Counties") 
} else 
{ print("There is no Statistical difference between the variances of hours worked by in Gwinnett and Fulton Counties")
}


```

We would like to visualize the previous results of our analysis.  Let us recall, that the averages of hours worked for RN's varies by states and counties.  Not only is there a statistical significant difference between the average, but the shape of the distribution is quite distinct, which is apparent by looking at the PDF and CDF plots.  

```{r, echo = FALSE}
## -- Plotting Section -- ##
# Filter data for cleaner plots (remove outliers)
# Boxplots
n <- 15

# 1) CNA hours worked by State
# a) 15 Most and Least

filtered_df <- df2 %>% group_by(State) %>% summarize(avg_CNA_hours = mean(Hrs_CNA, na.rm = TRUE)) %>% top_n(n, avg_CNA_hours) %>% inner_join(df, by = "State")
ggplot(data=filtered_df, aes(x = State, y = Hrs_CNA, fill = State)) +
  geom_violin(data = filtered_df, trim = FALSE, draw_quantiles = c(0.25, 0.5, 0.75)) + 
  geom_boxplot(data=filtered_df, aes(filtered_df$State, filtered_df$Hrs_CNA), width = 0.0075) + 
  labs(x = "State", y = "Hours Worked", title = "Box Plot: Hours worked via State for Certified Nurse Assistants (Top 15)") +
  scale_fill_manual(values = viridis::turbo(n = length(unique(filtered_df$State))+4)) +
  theme_linedraw()


# b) Bottom 15
filtered_df <- df2 %>% group_by(State) %>% summarize(avg_CNA_hours = mean(Hrs_CNA, na.rm = TRUE)) %>% top_n(-n, avg_CNA_hours) %>% inner_join(df, by = "State")
ggplot(data=filtered_df, aes(x = State, y = Hrs_CNA, fill = State)) +
  #geom_violin(data = filtered_df, trim = FALSE, draw_quantiles = c(0.25, 0.5, 0.75)) + 
  geom_boxplot(data=filtered_df, aes(filtered_df$State, filtered_df$Hrs_CNA), width = 0.75) + 
  labs(x = "State", y = "Hours Worked", title = "Box Plot: Hours worked via State for Certified Nurse Assistants (Bottom 15)") +
  scale_fill_manual(values = viridis::turbo(n = length(unique(filtered_df$State))+4)) +
  theme_linedraw()


# c) Histogram: FL vs GA
df_tmp <- df %>% filter(State == "GA" | State == "FL")
ggplot(data=df_tmp, aes(x = Hrs_CNA, fill = State)) +
  geom_histogram(data=df_tmp, aes(fill=df_tmp$State, x=df_tmp$Hrs_CNA, alpha = 0.5)) + 
  labs(x = "Hours worked", y = "Frequency", title = "Histogram: Hours worked by Certified Nurse Assistants - GA vs FL", fill = "State") +
  scale_fill_manual(values = turbo(n = length(unique(df_tmp$State))+4)) +
  theme_linedraw()


# d) PDF

ggplot(data=df_tmp, aes(x = df_tmp$Hrs_CNA, fill = State)) +
  geom_density(data=df_tmp, aes(fill=State, x=df_tmp$Hrs_CNA, alpha = 0.1)) + 
  labs(x = "Hours Worked", y = "Probability", title = "PDF: Hours worked by Certified Nurse Assistants", color = "State") +
  scale_fill_manual(values = viridis::turbo(n = length(unique(df_tmp$State))+4)) +
  theme_linedraw()


# e) CDF

ggplot(data=df_tmp, aes(x = df_tmp$Hrs_CNA, color = State)) +
  stat_ecdf(data=df_tmp, geom = "step", aes(color=State, x=df_tmp$Hrs_CNA)) + 
  labs(x = "Hours Worked", y = "Probability", title = "CDF: Hours worked by Certified Nurse Assistants", color = "State") +
  scale_color_manual(values = viridis::turbo(n = length(unique(df_tmp$State))+4)) +
  theme_linedraw()


# f) Histogram: Gwinnett vs Fulton
df_tmp <- df %>% filter(County_Name == "Fulton" | County_Name == "Gwinnett")
ggplot(data=df_tmp, aes(x = Hrs_CNA, fill = County_Name)) +
  geom_histogram(data=df_tmp, aes(fill=County_Name, x=Hrs_CNA, alpha = 0.5)) + 
  labs(x = "Hours worked", y = "Frequency", title = "Histogram: Hours worked by Certified Nurse Assistants - Fulton & Gwinnett Counties (GA)", fill = "County") +
  scale_fill_manual(values = turbo(n = length(unique(df_tmp$State))+4)) +
  theme_linedraw()



# g) PDF

ggplot(data=df_tmp, aes(x = df_tmp$Hrs_CNA, fill = County_Name)) +
  geom_density(data=df_tmp, aes(fill=County_Name, x=df_tmp$Hrs_CNA, alpha = 0.1)) + 
  labs(x = "Hours Worked", y = "Probability", title = "PDF: Hours worked by Certified Nurse Assistants", fill = "County") +
  scale_fill_manual(values = viridis::turbo(n = length(unique(df_tmp$County_Name))+4)) +
  theme_linedraw()


# h) CDF

ggplot(data=df_tmp, aes(x = df_tmp$Hrs_CNA, color = County_Name)) +
  stat_ecdf(data=df_tmp, geom = "step", aes(color=County_Name, x=df_tmp$Hrs_CNA)) + 
  labs(x = "Hours Worked", y = "Probability", title = "CDF: Hours worked by Certified Nurse Assistants", color = "County") +
  scale_color_manual(values = viridis::turbo(n = length(unique(df_tmp$County_Name))+4)) +
  theme_linedraw()


# 2) LPN hours worked by State
# a) Top 15

filtered_df <- df2 %>% group_by(State) %>% summarize(avg_LPN_hours = mean(Hrs_LPN, na.rm = TRUE)) %>% top_n(n, avg_LPN_hours) %>% inner_join(df, by = "State")
ggplot(data=filtered_df, aes(x = State, y = Hrs_LPN, fill = State)) +
  geom_violin(data = filtered_df, trim = FALSE, draw_quantiles = c(0.25, 0.5, 0.75)) + 
  geom_boxplot(data=filtered_df, aes(filtered_df$State, filtered_df$Hrs_LPN), width = 0.0075) + 
  labs(x = "State", y = "Hours Worked", title = "Box Plot: Hours worked via State for Licensed Practical Nurses (Top 15)") +
  scale_fill_manual(values = viridis::turbo(n = length(unique(filtered_df$State))+4)) +
  theme_linedraw()

# b) Bottom 15

filtered_df <- df2 %>% group_by(State) %>% summarize(avg_LPN_hours = mean(Hrs_LPN, na.rm = TRUE)) %>% top_n(-n, avg_LPN_hours) %>% inner_join(df, by = "State")
ggplot(data=filtered_df, aes(x = State, y = Hrs_LPN, fill = State)) +
  #geom_violin(data = filtered_df, trim = FALSE, draw_quantiles = c(0.25, 0.5, 0.75)) + 
  geom_boxplot(data=filtered_df, aes(filtered_df$State, filtered_df$Hrs_LPN), width = 0.75) + 
  labs(x = "State", y = "Hours Worked", title = "Box Plot: Hours worked via State for Licensed Practical Nurses (Bottom 15)") +
  scale_fill_manual(values = viridis::turbo(n = length(unique(filtered_df$State))+4)) +
  theme_linedraw()


# c) Histogram: FL vs GA

df_tmp <- df %>% filter(State == "GA" | State == "FL")
ggplot(data=df_tmp, aes(x = Hrs_LPN, fill = State)) +
  geom_histogram(data=df_tmp, aes(fill=State, x=Hrs_LPN, alpha = 0.5)) + 
  labs(x = "Hours worked", y = "Frequency", title = "Histogram: Hours worked by Licensed Practical Nurses - GA & FL", fill = "State") +
  scale_fill_manual(values = turbo(n = length(unique(df_tmp$State))+4)) +
  theme_linedraw()


# d) PDF

ggplot(data=df_tmp, aes(x = df_tmp$Hrs_LPN, fill = State)) +
  geom_density(data=df_tmp, aes(fill=df_tmp$State, x=df_tmp$Hrs_LPN, alpha = 0.1)) + 
  labs(x = "Hours Worked", y = "Probability", title = "PDF: Hours worked by Licensed Practical Nurses", color = "State") +
  scale_fill_manual(values = viridis::turbo(n = length(unique(df_tmp$State))+4)) +
  theme_linedraw()


# e) CDF

ggplot(data=df_tmp, aes(x = df_tmp$Hrs_LPN, color = df_tmp$State)) +
  stat_ecdf(data=df_tmp, geom = "step", aes(color=df_tmp$State, x=df_tmp$Hrs_LPN)) + 
  labs(x = "Hours Worked", y = "Probability", title = "CDF: Hours worked by Licensed Practical Nurses", color = "State") +
  scale_color_manual(values = viridis::turbo(n = length(unique(df_tmp$State))+4)) +
  theme_linedraw()


# f) Histogram: Filter by County for GA

df_tmp <- df %>% filter(State == "GA") %>% filter(County_Name == "Gwinnett" | County_Name == "Fulton")
ggplot(data=df_tmp, aes(x = Hrs_LPN, fill = County_Name)) +
  geom_histogram(data=df_tmp, aes(fill=County_Name, x=df_tmp$Hrs_LPN, alpha = 0.5)) + 
  labs(x = "Hours worked", y = "Frequency", title = "Histogram: Hours worked by Licensed Practical Nurses - Fulton & Gwinnett Counties (GA)", fill = "County") +
  scale_fill_manual(values = turbo(n = length(unique(df_tmp$State))+4)) +
  theme_linedraw()

# g) PDF

ggplot(data=df_tmp, aes(x = df_tmp$Hrs_LPN, fill = County_Name)) +
  geom_density(data=df_tmp, aes(fill=County_Name, x=df_tmp$Hrs_LPN, alpha = 0.1)) + 
  labs(x = "Hours Worked", y = "Probability", title = "PDF: Hours worked by Licensed Practical Nurses", fill = "County") +
  scale_fill_manual(values = viridis::turbo(n = length(unique(df_tmp$County_Name))+4)) +
  theme_linedraw()


# h) CDF

ggplot(data=df_tmp, aes(x = df_tmp$Hrs_LPN, color = County_Name)) +
  stat_ecdf(data=df_tmp, geom = "step", aes(color=County_Name, x=df_tmp$Hrs_LPN)) + 
  labs(x = "Hours Worked", y = "Probability", title = "CDF: Hours worked by Licensed Practical Nurses", color = "County") +
  scale_color_manual(values = viridis::turbo(n = length(unique(df_tmp$County_Name))+4)) +
  theme_linedraw()





# 3) RN hours worked by State
# a) Top 15

filtered_df <- df2 %>% group_by(State) %>% summarize(avg_RN_hours = mean(Hrs_RN, na.rm = TRUE)) %>% top_n(n, avg_RN_hours) %>% inner_join(df, by = "State")
ggplot(data=filtered_df, aes(x = State, y = Hrs_RN, fill = State)) +
  geom_violin(data = filtered_df, trim = FALSE, draw_quantiles = c(0.25, 0.5, 0.75)) + 
  geom_boxplot(data=filtered_df, aes(filtered_df$State, filtered_df$Hrs_RN), width = 0.0075) + 
  labs(x = "State", y = "Hours Worked", title = "Box Plot: Hours worked via State for Registered Nurses (Top 15)") +
  scale_fill_manual(values = viridis::turbo(n = length(unique(filtered_df$State))+4)) +
  theme_linedraw()

# b) Bottom 15

filtered_df <- df2 %>% group_by(State) %>% summarize(avg_RN_hours = mean(Hrs_RN, na.rm = TRUE)) %>% top_n(n, avg_RN_hours) %>% inner_join(df, by = "State")
ggplot(data=filtered_df, aes(x = State, y = Hrs_RN, fill = State)) +
  #geom_violin(data = filtered_df, trim = FALSE, draw_quantiles = c(0.25, 0.5, 0.75)) + 
  geom_boxplot(data=filtered_df, aes(filtered_df$State, filtered_df$Hrs_RN), width = 0.75) + 
  labs(x = "State", y = "Hours Worked", title = "Box Plot: Hours worked via State for Registered Nurses (Bottom 15)") +
  scale_fill_manual(values = viridis::turbo(n = length(unique(filtered_df$State))+4)) +
  theme_linedraw()

# c) Histogram: FL vs GA

df_tmp <- df %>% filter(State == "GA" | State == "FL")
ggplot(data=df_tmp, aes(x = Hrs_RN, fill = State)) +
  geom_histogram(data=df_tmp, aes(fill=df_tmp$State, x=df_tmp$Hrs_RN, alpha = 0.5)) + 
  labs(x = "Hours Worked", y = "Frequency", title = "Histogram: Hours worked by Registered Nurses - GA & FL", fill = "Gender") +
  scale_fill_manual(values = turbo(n = length(unique(df_tmp$State))+4)) +
  theme_linedraw()

# d) PDF

ggplot(data=df_tmp, aes(x = df_tmp$Hrs_RN, fill = df_tmp$State)) +
  geom_density(data=df_tmp, aes(fill=df_tmp$State, x=df_tmp$Hrs_RN, alpha = 0.1)) + 
  labs(x = "Hours Worked", y = "Probability", title = "PDF: Hours worked by Registered Nurses", fill = "State") +
  scale_fill_manual(values = viridis::turbo(n = length(unique(df_tmp$State))+4)) +
  theme_linedraw()


# e) CDF

ggplot(data=df_tmp, aes(x = df_tmp$Hrs_RN, color = df_tmp$State)) +
  stat_ecdf(data=df_tmp, geom = "step", aes(color=df_tmp$State, x=df_tmp$Hrs_RN)) + 
  labs(x = "Hours Worked", y = "Probability", title = "CDF: Hours worked by Registered Nurses", color = "State") +
  scale_color_manual(values = viridis::turbo(n = length(unique(df_tmp$State))+4)) +
  theme_linedraw()


# f) Histogram: Filter by County for GA

df_tmp <- df %>% filter(State == "GA") %>% filter(County_Name == "Gwinnett" | County_Name == "Fulton")
ggplot(data=df_tmp, aes(x = Hrs_RN, fill = County_Name)) +
  geom_histogram(data=df_tmp, aes(fill=County_Name, x=df_tmp$Hrs_RN, alpha = 0.5)) + 
  labs(x = "Hours Worked", y = "Frequency", title = "Histogram: Hours worked by Registered Nurses - Fulton & Gwinnett Counties (GA)", fill = "County") +
  scale_fill_manual(values = turbo(n = length(unique(df_tmp$State))+4)) +
  theme_linedraw()

# g) PDF

ggplot(data=df_tmp, aes(x = df_tmp$Hrs_RN, fill = County_Name)) +
  geom_density(data=df_tmp, aes(fill=County_Name, x=df_tmp$Hrs_RN, alpha = 0.1)) + 
  labs(x = "Hours Worked", y = "Probability", title = "PDF: Hours worked by Registered Nurses", fill = "County") +
  scale_fill_manual(values = viridis::turbo(n = length(unique(df_tmp$County_Name))+4)) +
  theme_linedraw()


# h) CDF

ggplot(data=df_tmp, aes(x = df_tmp$Hrs_RN, color = County_Name)) +
  stat_ecdf(data=df_tmp, geom = "step", aes(color=County_Name, x=df_tmp$Hrs_RN)) + 
  labs(x = "Hours Worked", y = "Probability", title = "CDF: Hours worked by Registered Nurses", color = "County") +
  scale_color_manual(values = viridis::turbo(n = length(unique(df_tmp$County_Name))+4)) +
  theme_linedraw()



```

Interestingly, we observe that a decent proportion of Registered Nurses work more hours in Gwinnett county compared to Fulton county.  We also observe Registered Nurses work more hours in the state of Florida compared to Georgia (on average) indicated by the CDF plot.  Our results focus mainly on geography (states and counties), but there could be demographic factors of the population that affect the hours worked by employees.

## Part 2
This section consists of SQL syntax embedded in R.  Similar results can be obtained by using an SQL programming language, but is included here for simplicity.

* Write a query to return the customer_name, product_name, and total_amount for each sale in the last 30 days.
* Write a query to find the total revenue generated by each product category in the last year. The output should include the product category and the total revenue for that category.
* Write a query to return all customers who made purchases in 2023 and are located in the "West" region.
* Write a query to display the total number of sales, total quantity sold, and total revenue for each customer. The result should include the customer_name, total sales, total quantity, and total revenue.
* Write a query to find the top 3 customers (by total revenue) in the year 2023.
* Write a query to rank products by their total sales quantity in 2023. The result should include the product_name, total quantity sold, and rank.
* Write a query that categorizes customers into "New" (if they signed up in the last 6 months) or "Existing" based on their sign_up_date. Include the customer_name, region, and category in the result.
* Write a query to return the month and year along with the total sales for each month for the last 12 months.
* Write a query to return the product categories that generated more than $50,000 in revenue during the last 6 months.
* Write a query to check for any sales where the total_amount doesn’t match the expected value (i.e., quantity * price).

```{r, echo = FALSE}

# PBJ Analysis pt 2

library(dplyr)

# Create sample data frames
sales <- data.frame(
  sales_id = 1:10,
  customer_id = c(1, 2, 3, 1, 2, 3, 1, 2, 3, 1),
  product_id = c(1, 2, 3, 2, 1, 3, 1, 2, 3, 1),
  sale_date = as.Date(c(Sys.Date() - 34 - years(1), Sys.Date() - 32 - years(1), Sys.Date() - 30, Sys.Date() - 24, Sys.Date() - 20, Sys.Date() - 16, Sys.Date() - 10, Sys.Date() - 8, Sys.Date() - 5, Sys.Date() - 2)),
  quantity = c(2, 3, 1, 4, 2, 3, 1, 2, 3, 1),
  total_amount = c(50, 75, 25, 100, 50, 75, 25, 50, 75, 25)
)

customers <- data.frame(
  customer_id = 1:3,
  customer_name = c("Alice", "Bob", "Charlie"),
  sales_region = c("North", "South", "West"),
  sign_up_date = as.Date(c("2023-12-01", "2023-12-15", "2024-01-01"))
)

products <- data.frame(
  product_id = 1:3,
  product_name = c("Product A", "Product B", "Product C"),
  category = c("Sports", "Clothing", "Home Goods"),
  price = c(10, 20, 15)
)

# Mutate the sales.total_amount to be the product of products.price * sales.quantity


# Export the data to csv files.
write.csv(sales, "ClipboardHealth_Sample_Sales.csv")
write.csv(customers, "ClipboardHealth_Sample_Customers.csv")
write.csv(products, "ClipboardHealth_Sample_Products.csv")



library(dplyr)
library(lubridate)
library(tidyr)

# Assuming your data is in a data frame named 'sales_data'
# ...

# 1. Sales in the last 30 days:

sales_last_30_days <- sales %>%
  inner_join(products, by = "product_id") %>%
  inner_join(customers, by = "customer_id") %>%
  filter(sale_date >= Sys.Date() - days(30)) %>%
  select(customer_name, product_name, total_amount)

# 2. Total revenue by product category within the last year:

product_category_revenue <- sales %>%
  inner_join(products, by = "product_id") %>%
  filter(sale_date >= Sys.Date() - years(1)) %>%
  group_by(category) %>%
  summarize(total_revenue = sum(total_amount))

# 3: Customers that made a purchase in 2023 in the 'West' region

west_region_customers <- sales %>%
  inner_join(customers, by = "customer_id") %>%
  filter(sale_date >= "2023-01-01" & sale_date <= "2023-12-31",
         sales_region == "West") %>%
  select(customer_name)

# 4: Total Sales, quantity, and revenue for each customer

customer_summary <- sales %>%
  inner_join(customers, by = "customer_id") %>%
  group_by(customer_name) %>%
  summarize(total_sales = sum(total_amount),
            total_quantity = sum(quantity))

# 5: Top 3 customers by total revenue in 2023 

top_customers <- sales %>%
  inner_join(customers, by = "customer_id") %>%
  filter(sale_date >= "2023-01-01" & sale_date <= "2023-12-31") %>%
  group_by(customer_name) %>%
  summarize(total_revenue = sum(total_amount)) %>%
  top_n(3, total_revenue)

# 6: Rank products by total sales quantity in 2023

product_rank <- sales %>%
  inner_join(products, by = "product_id") %>%
  filter(sale_date >= "2023-01-01" & sale_date <= "2023-12-31") %>%
  group_by(product_name) %>%
  summarize(total_quantity = sum(quantity)) %>%
  arrange(desc(total_quantity)) %>%
  mutate(rank = row_number())

# 7: Categorize customers as New or Existing:

customers_with_category <- customers %>%
  mutate(customer_category = case_when(
    sign_up_date >= Sys.Date() - months(6) ~ "New",
    TRUE ~ "Existing"
  ))

# 8: Total sales for each month in the last 12 months

monthly_sales <- sales %>%
  mutate(month_year = format(sale_date, "%Y-%m")) %>%
  group_by(month_year) %>%
  summarize(total_sales = sum(total_amount))

# 9: Product categories with revenue over 50,000

high_revenue_categories <- sales %>%
  inner_join(products, by = "product_id") %>%
  filter(sale_date >= Sys.Date() - months(6)) %>%
  group_by(category) %>%
  summarize(total_revenue = sum(total_amount)) %>%
  filter(total_revenue > 50000)

# 10: Check for mismatched total_amount

mismatched_sales <- sales %>%
  inner_join(products, by = "product_id") %>%
  filter(total_amount != quantity * price) %>% 
  select(sales_id, product_id, product_name, category, price, quantity, total_amount)


```