---
title: "Bakery Sales Mock Project"
author: "Anthony Morciglio"
date: "2024-07-25"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# Load libraries of the data:

library(ggplot2)
library(dplyr)
library(tidyverse)
library(readr)
library(skimr)
library(readxl)
library(viridis)
library(flextable)

options(warn = -1)

f1 <- "C:/Users/Anthony Morciglio/OneDrive/Coursera Google Data Analytics/Bakery Sales 2020 RW.csv"
df <- read.csv(f1)

```


## A Small Project to understand basic features in R

This mock project consists of data points with different features such as the following:

- Name of the item (Nombre).

- Cost of the item (Cuesta).

- Quantity of the item (Cantidad).

- Revenue of the items (Dinero) calculated by taking the product of quantity and cost.

## Look at the data in the form of a Table

```{r, echo = TRUE}

# Summarize key features of the data:
df
glimpse(df)
str(df)
summary(df)
skim_without_charts(df)

```

## Summarize Via Table

We can aggregate the data based on name type.  That is, group by name and summarize the total price and number of goods sold by using the group_by() and summarise() functions.

```{r, echo = TRUE}
TP <- df %>% group_by(Nombre) %>% summarise(Total_Price = sum(Cuesta * Cantidad), Total_Sold = sum(Cantidad))
TP %>% flextable()
```


## Slide with Graphing

```{r, echo = FALSE}
tmp <- data.frame(TP)
ggplot(data = tmp, aes(x = Nombre, y = Total_Price), fill = 'Nombre') + 
  geom_bar(stat = 'identity', aes(fill = tmp$Nombre)) + 
    labs(x = "Nombre", y = "Total Price", title = "Bar Chart: Total Price vs Nombre", fill = "Nombre") +
  scale_fill_manual(values = viridis(n = length(unique(TP$Nombre)))) + 
  theme_linedraw(base_size = 18)
                 
ggplot(data = tmp, aes(x = Nombre, y = Total_Sold), fill = 'Nombre') + 
  geom_bar(stat = 'identity', aes(fill = tmp$Nombre)) + 
    labs(x = "Nombre", y = "Total Sold", title = "Bar Chart: Total Sold vs Nombre", fill = "Nombre") +
  scale_fill_manual(values = viridis(n = length(unique(TP$Nombre)))) + 
  theme_linedraw(base_size = 18)

```


## Takeaways
Cupcakes are the most popular item at the bakery.  Over 250 Cupcakes were sold at the local bakery this week compared to a less popular item, the Muffin, which had 54 of them sold.  Although Pies and Muffins had similar quantities (cuantidad) sold, Pies brought much more profit since it cost $5 per each.