---
title: "Why Choose R?"
author: "Anthony Morciglio"
date: "2024-06-05"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(palmerpenguins)
library(readr)
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r penguins, echo = TRUE}
summary(penguins)
penguins <- penguins %>% drop_na()
colnames(penguins)[7] <- "gender"
```

## Slide with Plot

```{r penguins plots}
ggplot(data = penguins, aes(x = flipper_length_mm, fill = species)) + 
  geom_density(data = penguins, aes(x = penguins$flipper_length_mm, fill = penguins$species, alpha = 0.2)) + 
  theme_linedraw()

ncat = length(unique(penguins$species))
col_palette <- rev(RColorBrewer::brewer.pal(ncat, "YlGn"))


ggplot(data = penguins, aes(x = flipper_length_mm, group = species, fill = species)) +
  geom_density(position = "fill", adjust = 1.5) +
  labs(x = "Flipper Length", y = "Density", title = "Density Plot: Flipper Length Categorized by Species", fill = "UserType") +
  scale_fill_manual(values = col_palette) + 
  theme_linedraw(base_size = 18) 

```



