---
title: "Entropy of the Normal Distribution"
author: "Anthony Morciglio"
date: "10/03/2025"
output: html_document
---

```{r setup, include=FALSE, warning = FALSE, message = FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Entropy of the Normal Distribution

The entropy of a Normal distribution (N(\mu, \sigma^2)) is given by:

[
H(Y) = \tfrac{1}{2} \ln(2 \pi e \sigma^2).
]

This depends only on the variance, not the average

### Example: Normal(0, 4)

Let’s compute the entropy for (\mu = 0, \sigma^2 = 4).

```{r}
mu <- 0
sigma <- 2
entropy_nats <- 0.5 * log(2 * pi * exp(1) * sigma^2)
entropy_bits <- entropy_nats / log(2)

entropy_nats
entropy_bits
```

* Entropy ≈ 2.11 nats
* Entropy ≈ 3.04 bits

### Entropy as a Function of Variance

We now compute entropy for a range of (\sigma) values to see how it grows.

```{r}
sigma_vals <- seq(0.5, 5, by = 0.5)
entropy_vals <- 0.5 * log(2 * pi * exp(1) * sigma_vals^2)

df <- data.frame(sigma = sigma_vals, entropy = entropy_vals)

ggplot(df, aes(x = sigma, y = entropy)) +
  geom_line(color = "blue", size = 1.2) +
  geom_point(color = "black", size = 2) +
  labs(title = "Entropy of the Normal Distribution",
       x = "Standard Deviation (σ)",
       y = "Entropy (nats)") +
  theme_minimal()
```

### Monte Carlo Verification

We can approximate entropy numerically by simulating from the distribution and evaluating (-\mathbb{E}[\ln f(Y)]).

```{r}
set.seed(123)
n <- 1e5
y <- rnorm(n, mean = mu, sd = sigma)
f_y <- dnorm(y, mean = mu, sd = sigma)
entropy_mc <- -mean(log(f_y))

entropy_mc
```

This Monte Carlo estimate should be close to the theoretical entropy.

### Boxplot of Entropy Estimates

If we repeat the Monte Carlo approximation several times, we can see the variability in estimates.

```{r}
mc_reps <- replicate(20, {
  y <- rnorm(n, mean = mu, sd = sigma)
  f_y <- dnorm(y, mean = mu, sd = sigma)
  -mean(log(f_y))
})

boxplot(mc_reps, horizontal = TRUE,
        main = "Monte Carlo Estimates of Entropy",
        xlab = "Entropy (nats)",
        col = "orange")
abline(v = entropy_nats, col = "black", lwd = 2, lty = 2)
```

The red dashed line shows the theoretical entropy, and the Monte Carlo estimates cluster closely around it.

```
```
