---
title: "Exponential Distribution: Quantile and CDF Functions"
author: "Anthony Morciglio"
date: "2025-08-23"
output: html_document
---


## Applying the Quantile Function to Visualize the Value corresponding to the Probability of an Exponential Distribution

This R script provides a hands-on illustration of the exponential distribution's Cumulative Distribution Function (CDF) and its inverse, the Quantile Function. The two plots produced by the script serve as a visual companion to the mathematical derivations. The first plot shows the traditional CDF, mapping a value on the x-axis to its cumulative probability on the y-axis. The second plot then inverts this relationship, using the quantile function to map a probability on the x-axis to the corresponding value on the y-axis. Together, these visualizations clearly demonstrate how the quantile function is used to find specific points in the distribution, such as the median or other percentiles, given a desired probability.

```{r setup, warning = FALSE, message = FALSE, echo = FALSE
knitr::opts_chunk$set(echo = FALSE)
}
```

```{r setup}
# This R script provides functions to calculate and plot the
# exponential distribution's Cumulative Distribution Function (CDF)
# and its inverse, the Quantile Function.

# Required Library:
# ggplot2 for creating high-quality, professional plots.
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  install.packages("ggplot2")
}
library(ggplot2)

# -----------------------------------------------------------------------------
# 1. Define the Quantile and CDF Functions
# -----------------------------------------------------------------------------
# The Quantile Function (inverse of the CDF)
# Q(p) = -1/lambda * ln(1 - p)
# @param p The probability or quantile (from 0 to 1).
# @param lambda The rate parameter of the exponential distribution.
# @return The value x corresponding to the probability p.
quantile_exponential <- function(p, lambda) {
  return(-1 / lambda * log(1 - p))
}

# The Cumulative Distribution Function (CDF)
# F(x) = 1 - e^(-lambda * x)
# @param x The value to evaluate the CDF at.
# @param lambda The rate parameter.
# @return The cumulative probability.
cdf_exponential <- function(x, lambda) {
  return(1 - exp(-lambda * x))
}

# -----------------------------------------------------------------------------
# 2. Demonstrate the Inverse Relationship
# -----------------------------------------------------------------------------
# Set a sample lambda value.
lambda_val <- 0.5

# Let's find the value for the 50th percentile (the median).
p_val <- 0.50
x_val <- quantile_exponential(p_val, lambda_val)

cat(sprintf("Using lambda = %.2f\n", lambda_val))
cat("--------------------------------\n")
cat(sprintf("The value for the %.0f%% percentile is x = %.4f\n", p_val * 100, x_val))

# Now, we can check if this value maps back to the original probability using the CDF.
check_p <- cdf_exponential(x_val, lambda_val)
cat(sprintf("Checking the CDF at x = %.4f gives a probability of p = %.4f\n", x_val, check_p))
cat("--------------------------------\n")


# Let's find the value for the 95th percentile.
p_val_95 <- 0.95
x_val_95 <- quantile_exponential(p_val_95, lambda_val)
cat(sprintf("The value for the %.0f%% percentile is x = %.4f\n", p_val_95 * 100, x_val_95))

# -----------------------------------------------------------------------------
# 3. Create a Visualization to Illustrate the Inverse Relationship
# -----------------------------------------------------------------------------
# Create a data frame for plotting the CDF.
x_range <- seq(0.01, 8.01, length.out = 500)
cdf_data <- data.frame(
  x = x_range,
  p = cdf_exponential(x_range, lambda_val)
)

# Plot the CDF.
cdf_plot <- ggplot(cdf_data, aes(x = x, y = p)) +
  # Add the main CDF curve.
  geom_line(color = "blue", linewidth = 1) +
  # Add horizontal line from the 50th percentile.
  geom_hline(yintercept = p_val, linewidth = 1.25, linetype = "dashed", color = "black") +
  # Add vertical line from the 50th percentile to the x-axis.
  geom_vline(xintercept = x_val, linewidth = 1.25, linetype = "dashed", color = "black") +
  # Mark the intersection point.
  geom_point(aes(x = x_val, y = p_val), color = "orange", size = 3) +
  # Add labels and a title.
  labs(
    title = "Exponential Distribution CDF: Value from Probability",
    subtitle = bquote(lambda == .(lambda_val)),
    x = "x (Value)",
    y = "p (Probability)"
  ) +
  # Set the theme and customize aesthetics.
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

print(cdf_plot)


# -----------------------------------------------------------------------------
# 4. Create a Visualization to Illustrate the Quantile Function
# -----------------------------------------------------------------------------
# Create a range of probabilities for the plot.
p_range <- seq(0.001, 0.999, length.out = 500)

# Create a data frame for plotting the quantile function.
quantile_data <- data.frame(
  p = p_range,
  x = quantile_exponential(p_range, lambda_val)
)

# Plot the Quantile Function.
quantile_plot <- ggplot(quantile_data, aes(x = p, y = x)) +
  # Add the main Quantile Function curve.
  geom_line(color = "red", linewidth = 1) +
  # Add vertical line from the 50th percentile.
  geom_vline(xintercept = p_val, linetype = "dashed", linewidth = 1.25, color = "black") +
  # Add horizontal line from the 50th percentile to the y-axis.
  geom_hline(yintercept = x_val, linetype = "dashed", linewidth = 1.25, color = "black") +
  # Mark the intersection point.
  geom_point(aes(x = p_val, y = x_val), color = "orange", size = 3) +
  # Add labels and a title.
  labs(
    title = "Exponential Distribution Quantile Function: Probability to Value",
    subtitle = bquote(lambda == .(lambda_val)),
    x = "p (Probability)",
    y = "x (Value)"
  ) +
  # Set the theme and customize aesthetics.
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

print(quantile_plot)
```


